A comparison of different methods for calculating tangent-stiffness matrices in a massively parallel computational peridynamics code.

### 2013-11-25 transition.tex
Made springer edits CMAME comptible

%% if you use PostScript figures in your article
%% use the graphics package for simple commands
%%\usepackage{graphics}
%% or use the graphicx package for more complicated commands
%%\usepackage{graphicx}
%% or use the epsfig package if you prefer to use the old commands
%% \usepackage{epsfig}
%%\usepackage{color}
%%\usepackage{tkz-base}
\usepackage{pgfplots}
\usepackage{siunitx}
%\usepackage{hyperref}

\usepackage{array}

%% The amssymb package provides various useful mathematical symbols
\usepackage{amssymb}

@@

\begin{abstract} %% Text of abstract Shown is a retrospective comparative study of tangent-stiffness
In order to maintain the quadratic convergence properties of the first-order Newton's method in quasi-static nonlinear analysis of solid structures it is crucial to obtain accurate, algorithmically consistent tangent-stiffness matrices. For an extremely small class of nonlinear material models, these consistent tangent-stiffness operators can be derived analytically; however, most often in practice, they are found through numerical approximation of derivatives.  This paper compares four methods of numerical derivative calculation: automatic differentiation, complex-step, forward finite difference, and central finite difference, in the context of tangent-stiffness matrix calculation in a massively parallel computational peridynamics code.  The methods are compared through \emph{in situ} profiling of the code for accuracy, speed, efficiency, and parallel scalability.  The research provides data that can serve as practical guide for code developers and analysts faced with choosing which method best suit the needs of their application code.  Additionally, motivated by the reproducible research movement, all the of the code, examples, and workflow to regenerate the data and figures in this paper are provided open-source.
\end{abstract}

@@

\section{Introduction}
%\subsection{Motivation}
\label{sec:intro}

In order to maintain the quadratic convergence properties of the first-order Newton's method \cite{belytschko1999nonlinear} \cite[Ch.~13]{young2009} in quasi-static nonlinear analysis of solid structures it is crucial to obtain accurate, algorithmically consistent tangent-stiffness matrices. For an extremely small class of nonlinear material models, these consistent tangent-stiffness operators can be derived analytically; however, most often in practice, they are found through numerical approximation of derivatives.

A goal of this study was to develop and evaluate a new, accurate, and practical method for calculating tangent-stiffness matrices against established methods.  This new method is based on a complex number Taylor series expansion and referred to as CTSE or the ``complex-step'' method.  The
distinction of `accurate' is defined by comparison of the new method to popular finite differencing techniques used for computing tangent-stiffness matrices and with the exact algorithmically consistent derivatives computed with \emph{automatic differentiation}.

@@

The scope of the application component of the study was limited to a single material model implemented in the computational peridynamics code, \textit{Peridigm} \cite{peridigm}. Identifying a specific application provided a practical framework for implementing the new method and solving engineering problems to generate the data needed to compare the methods. In particular, \emph{Peridigm} was chosen because it combined several helpful characteristics: the utilization of Newton's method and tangent-stiffness matrices for solving nonlinear quasi-static problems, prior inclusion of finite-difference, central-difference, and automatic-differentiation methods needed for comparison to the new complex-step method implementation, and being an agile-components code that makes use of distributed computing data structures via Trilinos \cite{trilinos} for efficient parallelization on large clusters.

The aim of this paper is not only to introduce the new complex-step method in the context of evaluating tangent-stiffness matrices, but to serve as a review for other differentiation techniques and as a general reference for solving non-linear systems with Newton's method. After presenting background information on the underlying methods to be discussed, presented in this paper are: a description of tangent-stiffness matrices and how to use them to solve non-linear systems, detailed directions for producing tangent-stiffness matrices with each the methods identified above, a description and justification of the new complex-step method for calculating tangent-stiffness matrices, a description of implementing complex-step in the \emph{Peridigm} software, a description of the quantities-of-interest used to rank the methods, a presentation and analysis of the results of the comparative study, and conclusions and thoughts on potential future work. Finally, in the interest of replicable research, the reader is referred to the corresponding author's website for C++ source-code and data necessary to reproduce the results presented here. Additionally, included on the website is a C++ library with classes for solving non-linear systems using the techniques discussed here, as well as two validation and verification example problems which show how to use the library.

@@

This section contains background information on the differentiation techniques underlying the tangent-stiffness matrix calculation methods compared in the study. Since first-order finite-difference techniques are considered well known, they are not described here. Instead the reader is referred to \cite[Chap. 4.1.3]{chapra2010}.

\subsubsection{The ``complex-step'' method.}  
\label{sec:CSmethod}
It is possible to approximate derivatives quite accurately with a technique based on a complex variable Taylor series expansion of a function.  This method was first described by Lyness and Moler \cite{lyness1967numerical, lyness1968differentiation} and has more recently been rediscovered \cite{squire1998using, voorhees2011complex, al2010complex} for use in engineering analysis.  The basic idea is that a model parameter can be be made to be complex and expanded in a Taylor series about a small perturbation in the imaginary plane by some arbitrary value $h$ as follows
%
\begin{equation}
f ( x + i h ) = f (x) + \frac{\partial f(x)}{\partial x} \frac{i h}{1!} + \frac{\partial f^2(x)}{\partial x^2} \frac{(i h)^2}{2!} + \frac{\partial f^3(x)}{\partial x_n^3} \frac{(i h)^3}{3!} + \cdots
\label{eqn:complexTaylor}
\end{equation}
%
Now we can take the imaginary part of both sides of equation (\ref{eqn:complexTaylor}) and solve for the first derivative ignoring terms of $\mathcal{O}\left ( h^2 \right)$ to yield an the estimate
%
\begin{equation}
 \frac{\partial f( x )}{\partial x} \approx \frac{\mbox{Im} \left( f (x + i h) \right)}{h}.
\label{eqn:complexFirstDeriv}
\end{equation}
%
This shows that an estimate of the first derivative of a function can be made by only utilizing one functional evaluation of a perturbed model parameter up the imaginary axis.  The step-size $h$ is arbitrary and can be made as small as practical (even to machine precision without the dangers of roundoff error as in finite-differences) to yield an accurate estimate of this derivative.  The only disadvantage of this technique is the necessity of requiring the functions to accept complex numbers as arguments. Appendix~\ref{sec:appendixA} includes a simple example that illustrates the use of the complex-step method for derivative calculation of a function.

@@

Automatic-differentiation (AD) is a computerized method for computing exact derivatives based the chain-rule from calculus. AD takes advantage of the fact that any mathematical function executed on a computer, no matter how complicated, is a ``composition of simple operations'' (add, multiply, power, transcendental and the like) each having known analytical derivatives \cite{ref-sacado-presentation}. For reference, the AD implementation used in the study is part of the \emph{Sacado} package from the \emph{Trilinos} agile component libraries developed at Sandia National Laboratories \cite{ref-Sacado}.

The way an AD system works is by first evaluating the innermost function of the composition, then presenting that function's output as input to the next level function until all levels are complete, in a way no different from how a normal computer program or human would evaluate composition functions.  AD departs in that as each nested function is evaluated, the function's partial derivative with respect to the designated variables of the given input is also calculated.  This is possible because the elementary math functions are hard-coded into the AD source-code along with their analytical derivatives, and linked by special instructions, so that when the elementary math functions are called upon for computation, their partial derivatives may be computed and stored in a sequence. The AD system then multiplies the final sequence of partial derivatives together to produce the exact equivalent to taking a partial derivative of the corresponding composition function with respect to a designated variable at a particular value. It is obvious, but bears mentioning, that the AD system could simply store one value for partial derivatives, modifying it as appropriate for every function evaluation rather than keeping a sequence.  In the literature, the AD scheme described here is called \emph{forward automatic-differentiation}. Only forward AD will be covered here since it is the implementation used in \emph{Sacado} and therefore this study; however, the reader is referred to the introduction section of \cite{ref-AD-methods} and its citations for further information on AD, particularly \cite{ref-on-AD} which is foundational. Appendix~\ref{sec:appendixB} includes an example that walks the reader through the process a computer uses to compute derivatives via AD.  Some things to note about AD are that no approximation of derivatives is being made because the analytical forms of the partials of the elementary math functions are defined alongside them. The accuracy of AD is then limited by the precision of the AD system's definition of the elementary math functions and their partial derivatives .

@@

In solid and structural computational mechanics the tangent-stiffness matrix is a linearization operator that describes the stiffness of a system in response to small displacements imposed upon the current configuration of the system.  Mathematically speaking, the tangent-stiffness represents the gradient of a high-dimensional energy surface that ``points'' in the minimum direction. While we've adopted the terminology of solid mechanics, it's important to note that these operators appear in other physical settings and are known by other names, e.g. a \emph{transmissibility matrix} in the context of a Poisson problem or, more generally, a \emph{Jacobian matrix} in mathematical optimization.

While, for simplicity, the examples shown in the sequel refer only to linear problems, tangent-stiffness matrices generally arise in the context of non-linear analysis where Newton's method or quasi-Newton's method's are used in the minimization of some residual function.  In this context, the tangent-stiffness matrix can be thought of as the linearization of the system about about a particular configuration.  It has been shown \cite{hughes1978consistent, hughes1978unconditionally} that in order to preserve the quadratic convergence properties of the Newton's methods, it is necessary that this linearization is carried out in a manner that is \emph{consistent} with the algorithmic constraint equations used when computing the internal forces that arise due to deformations.  An example of these constraint equations would be the Kuhn-Tucker conditions \cite{simo1998} that are used in integration of a flow rule for plasticity modeling.  If the continuum tangent-moduli are used in place of the \emph{consistent} or \emph{algorithmic tangent moduli} in the solution of a non-linear solid mechanics problem, convergence may still be achieved, but not in the quadratic manner that makes Newton's method so attractive for this class of problems.  Again, in the setting of non-linear solid mechanics, there is a very small class of material model algorithms in which consistent tangent-stiffness operators can be derived analytically.  Perfect plasticity and plasticity with isotropic hardening used in combination with a general nearest-point projection or \emph{radial return algorithm} are a few examples.  Therefore, when these models (and certainly more complex ones) are implemented into a general purpose computational mechanics code, the tangent-stiffness operators are typically defined via a numerical approximation.

In detail, a tangent-stiffness matrix can be described as a collection of the first-order partial derivatives of a vector valued function with respect to each independent vector component of the function for a given vector.   A general mathematical formula for a tangent-stiffness matrix, $K_{ij}$,
can be expressed in indicial notation as 
%
\begin{equation} 
  K_{ij} = \left. \frac{\partial F_i}{\partial X_j}\right|_{\vec{X}_0},
\end{equation}
%
where $F_i$ is the $i^{\mbox{th}}$ component of the vector valued function, $X_j$ is the $j^{\mbox{th}}$ component of the vector argument to $F$, and a particular value of the vector is chosen to linearize about $\vec{X}_0$. One then evaluates the expression for each combination $(i, j)$ corresponding to a (row, column) location in the tangent-stiffness matrix. By inspection, the elements of a tangent-stiffness matrix can be estimated using any of the complex-step, AD, or common finite-difference techniques for functions $F:R^1 \rightarrow R^1$, since taking partial derivatives entails holding all but a single independent vector component of the function constant, and each element of the function can be evaluated independently of the others. \\

@@

The \emph{forward-difference} (FD) formula for calculating a tangent-stiffness matrix in the setting of a solid structural mechanics problem is
%
\begin{equation} 
  K_{ij} = \frac{F_i^{int}(\vec{u} + h \hat{e}_j) - F_i^{int}(\vec{u})}{h},
\end{equation}
%
Where $K_{ij}$ is the indicial notation representation of an element of the tangent-stiffness matrix at row $i$, column $j$. The first of the two terms in the numerator is the internal force, $F_i^{int}$ evaluated in the current deformed configuration, $\vec{u}$,  plus a small perturbation $h$ in the direction $\hat{e}_j$ where $\hat{e}_j$ represents a unit-vector corresponding to the $j^{\mbox{th}}$ degree-of-freedom. The second of the two terms is $F_i^{int}$ evaluated in the current configuration. The denominator is the magnitude of perturbation called \emph{probe-distance}.

The \emph{central-difference} (CD) formula for calculating a tangent-stiffness matrix is
%
\begin{equation} 
  K_{ij} = \frac{F_i^{int}(\vec{u} + h \hat{e}_j) - F_i^{int}(\vec{u} - h \hat{e}_j)}{2 h}.
\end{equation}
%
Together, the FD and CD methods, are sometimes termed \emph{finite-difference probing} techniques for tangent-stiffness calculation in literature \cite{ref-Adaggio}.

The complex-step (CS) formula for calculating a tangent-stiffness matrix is
%
\begin{equation} 
  K_{ij} = \frac{\mbox{Im}(F_i^{int}(\vec{u} + i h \hat{e}_j))}{h},
\end{equation}
%
where the internal force, $F^{int}$ is now treated as function of complex vectors,  $\vec{u}$. The small perturbation, $h$, is now carried out on the imaginary axis and only the real coefficient to the imaginary part of the vector returned by $F^{int}$ is kept, hence the $\mbox{Im}(\cdot)$ operation. As highlighted in Section~\ref{sec:CSmethod}, notice there is no subtraction operation taking place; therefore $h$ can be made very small to yield highly accurate derivatives.

@@

\section{Description of analysis tools and approach.} 
%
\subsection{Peridigm and  peridynamics.}
%
Comparisons of the different tangent-stiffness calculation techniques were carried out through code-development and \emph{in situ} instrumentation of the computational peridynamics code \emph{Peridigm} \cite{peridigm}. \emph{Peridigm} is distributed by Sandia National Laboratories as its primary open-source computational peridynamics code. It is a massively-parallel simulation code for implicit and explicit multi-physics simulations primarily used for solid mechanics and material failure. \emph{Peridigm} is a C++ code utilizing agile software components from Sandia's  \emph{Trilinos} project \cite{trilinos}. It's important to note that there is no specialization of the tangent-stiffness calculation methods described previously to this particular code or more generally to a computational peridynamics approach and the analysis carried out in this study should provide insight into the speed and accuracy of these methods when implemented into other computational mechanics analysis tools as well.  \emph{Peridigm} was chosen because of the author's relationship with the code development project and primarily because the FD, CD, and AD methods for tangent-stiffness calculation where already implemented in the code, leaving only the CS method to for development. However, because peridynamics is a nonlocal theory, the tangent-stiffness matrices have a much higher bandwidth (i.e. less sparsity) than traditional finite-element tangent-stiffness computations and the time required to construct the tangent-stiffness is a majority of the total computation time in any quasi-static or implicit dynamics analysis; therefore, the \emph{Peridigm} development team has an interest in profiling the performance of the tangent-stiffness computation methods.

Briefly, peridynamics \cite{silling2000ret, silling:psa, silling2010peridynamic} is a nonlocal reformulation of the partial-differential equations that provide the statement of momentum balance in classical continuum mechanics. Its primary goal is the avoid the use of spatial derivatives in the balance equations or constitutive laws such that discontinuous displacements, i.e. cracks, are mathematically consistent with the governing equations. It has demonstrated great promise in modeling problems with massive pervasive failure \cite{littlewood2010}, interesting phenomenon such as dynamic crack branching \cite{ha2010sod}, and is being extended to multiphyics phenomenon \cite{bobaru2011peridynamic,katiyar2013} and coupling to molecular simulations \cite{seleson2009peridynamics}.

\subsubsection{Implementing the complex-step method in Peridigm} 
%
The CS method was implemented in \emph{Perdigm} in a manner that follows closely the implementation of the FD technique.  However, because of \emph{Peridigm}'s reliance on \emph{Trilinos} and specifically the \emph{Epetra} package, special steps had to be followed to allow the use of complex number data types.  This is because \emph{Epetra} vectors, which can be thought of as distributed memory parallel cousins to standard C++ STL vectors, are hard coded to be of type {\tt double} and can not be simply be declared as having a complex data type. Therefore, program methods that were used in the course of applying the FD method were copied, renamed and re-written to follow the CS formula, which involved changing the data type of intermediate variables which were locally scoped and dynamically casting persistent memory variables to complex data types where necessary. We made no attempt to ``tune'' the CS code in any way for performance, we attempted to implement the algorithm is such as way that it was in one-to-one correspondence with the FD and CD methods; however, it must be noted that the overhead associated with the dynamic casts in the CS method likely hurt its overall performance and this issue could be alleviated by using the next generation templated \emph{TPetra} vectors available in \emph{Trilinos} that are capable of taking an explicit instantiation of any data type including complex.  There where additional modifications for the purpose instrumenting the code to collect comparative data and are not specifically part of CS or other methods. The source code patch to \emph{Peridigm} for implementing CS can be found on the corresponding author's website.

\emph{Peridigm} was primarily written by expert computational scientists; therefore, many advanced techniques in C++ such as multiple virtual inheritance, pointer arithmetic, and distributed data structures were used in the basic code and by necessity in the modified code.  To see CS and other methods in a more narrow context, it is advised that readers also take a look at the simple examples provided on the corresponding author's website.

>>>>

%% if you use PostScript figures in your article
%% use the graphics package for simple commands
%%\usepackage{graphics}
%% or use the graphicx package for more complicated commands
%%\usepackage{graphicx}
%% or use the epsfig package if you prefer to use the old commands
%% \usepackage{epsfig}
%%\usepackage{color}
%%\usepackage{tkz-base}
\usepackage{pgfplots}
\usepackage{siunitx}
\usepackage{graphicx}
\usepackage{amsmath}
\usepackage{booktabs}
\usepackage{siunitx}
\usepackage{array}
%% The amssymb package provides various useful mathematical symbols
\usepackage{amssymb}

@@

\begin{abstract} %% Text of abstract Shown is a retrospective comparative study of tangent-stiffness
In order to maintain the quadratic convergence properties of Newton's method in quasi-static nonlinear analysis of solid structures it is crucial to obtain accurate, algorithmically consistent tangent-stiffness matrices. A goal of the study described in this paper was to establish the suitability of an under-explored method for numerical computation of tangent-stiffness operators, referred to as ``complex-step'', and compare the new method with other techniques for numerical derivative calculation: automatic differentiation, forward finite-difference, and central finite-difference. The complex-step method was newly implemented in a massively parallel computational peridynamics code for the purpose of this comparison. The methods were compared through in situ profiling of the code for accuracy, speed, efficiency, and parallel scalability. The research provides data that can serve as practical guide for code developers and analysts faced with choosing which method best suits the needs of their application code. Additionally, motivated by the reproducible research movement, all the of the code, examples, and workflow to regenerate the data and figures in this paper are provided as open source.

\end{abstract}

@@

\section{Introduction}
%\subsection{Motivation}
\label{sec:intro} 
In order to maintain the quadratic convergence properties of the first-order Newton's method \cite{belytschko1999nonlinear} \cite[Ch.~13]{young2009} in quasi-static nonlinear analysis of solid structures it is crucial to obtain accurate, algorithmically consistent tangent-stiffness matrices. For an extremely small class of nonlinear material models, these consistent tangent-stiffness operators can be derived analytically; however, most often in practice, they are found through numerical approximation of derivatives.

A goal of this study was to develop and evaluate a new, accurate, and practical method for calculating tangent-stiffness matrices against established methods.  This new method is based on a complex number Taylor series expansion and referred to as CTSE or the ``complex-step'' method.  The
distinction of ``accurate'' is defined by comparison of the new method to popular finite differencing techniques used for computing tangent-stiffness matrices and with the exact to machine precision algorithmically consistent derivatives computed with \emph{automatic differentiation}.

@@

The scope of the application component of the study was limited to a single material model, an elastic peridynamic solid, implemented in the computational peridynamics code, \textit{Peridigm} \cite{peridigm}. Identifying a specific application provided a practical framework for implementing the new method and solving engineering problems to generate the data needed to compare the methods. In particular, \emph{Peridigm} was chosen because it combined several helpful characteristics: the utilization of Newton's method and tangent-stiffness matrices for solving nonlinear quasi-static problems, prior inclusion of finite-difference, central-difference, and automatic-differentiation methods needed for comparison to the new complex-step method implementation, and being an agile-components code that makes use of distributed computing data structures via Trilinos \cite{trilinos} for efficient parallelization on large clusters. Agile-components is a software development term meaning to structure a software project for continuous development to respond to evolving specifications by making much use of techniques from object-oriented programming and reusing proven pre-existing code to reduce development time and increase reliability. The modularity of \emph{Peridigm} allowed the adding of new features, such as required by the study.

The aim of this paper is not only to introduce the new complex-step method in the context of evaluating tangent-stiffness matrices, but to serve as a review for other differentiation techniques useful for solving non-linear systems with Newton's method. After presenting background information on the underlying methods to be discussed, presented in this paper are: A) a description of tangent-stiffness matrices, B) detailed directions for producing tangent-stiffness matrices with each the methods identified above, C) a description and justification of the new complex-step method for calculating tangent-stiffness matrices, D) a description of implementing complex-step in the \emph{Peridigm} software, E) a description of the quantities-of-interest used to rank the methods, F) a presentation and analysis of the results of the comparative study, and G) conclusions and thoughts on potential future work. Finally, in the interest of replicable research, the reader is referred to the corresponding author's website for \emph{Peridigm} C++ source-code and data necessary to reproduce the results presented here. Additionally, included on the website is an example serial C++ library with classes for solving non-linear systems using the techniques discussed here, as well as two validation and verification example problems which show how to use the library. The purpose of the serial C++ library is to demonstrate in a less complex manner than in the \emph{Peridigm} code how one would use automatic-differentiation or complex-step to better encourage their use and discussion. This open source resource will be referred to as the \emph{paper repository} in what follows.

@@

This section contains background information on the differentiation techniques underlying the tangent-stiffness matrix calculation methods compared in the study. Since the derivation of first-order finite-difference techniques from Taylor series expansion is considered well known, it is not described here. Instead the reader is referred to \cite[Chap. 4.1.3]{chapra2010}.

\subsubsection{The ``complex-step'' method.}  
\label{sec:CSmethod}
It is possible to approximate derivatives quite accurately with a technique based on a complex variable Taylor series expansion of a function.  This method was first described by Lyness and Moler \cite{lyness1967numerical,lyness1968differentiation} and has more recently been rediscovered \cite{squire1998using,voorhees2011complex,al2010complex,martins2000automated} for use in engineering analysis.  The basic idea is that a model parameter can be be made complex and expanded in a Taylor series about a small perturbation along the imaginary axis by some arbitrary value $h$ as follows
%
\begin{align}
f ( x + i h ) =& f(x) + \frac{\partial f(x)}{\partial x} \frac{i h}{1!}  \notag \\
               &+ \frac{\partial f^2(x)}{\partial x^2} \frac{(i h)^2}{2!} + \frac{\partial f^3(x)}{\partial x^3} \frac{(i h)^3}{3!} + \cdots,
\label{eqn:complexTaylor}
\end{align}
%
Taking the imaginary part of both sides of equation (\ref{eqn:complexTaylor}) and solving for the first derivative and ignoring terms of $\mathcal{O}\left ( h^2 \right)$ yields an estimate
%
\begin{equation}
 \frac{\partial f( x )}{\partial x} \approx \frac{\mbox{Im} \left( f (x + i h) \right)}{h}.
\label{eqn:complexFirstDeriv}
\end{equation}
%
Thus, an estimate of the first derivative of a function can be made by only utilizing one functional evaluation of a perturbed model parameter along the imaginary axis.  The step-size $h$ is arbitrary and can be made as small as practical (even to machine precision without the dangers of roundoff error as in finite-differences) to yield an accurate estimate of this derivative.  The only disadvantage of this technique is the necessity of requiring the functions to accept complex numbers as arguments. Appendix~\ref{sec:appendixA} includes a simple example that illustrates the use of the complex-step method for the derivative calculation of a function.

@@

Automatic-differentiation (AD) is a computerized method for computing exact derivatives based the chain-rule from calculus. AD takes advantage of the fact that any mathematical function executed on a computer, no matter how complicated, is a ``composition of simple operations'' (add, multiply, power, transcendental and the like) each having known analytical derivatives \cite{ref-sacado-presentation}. For reference, the AD implementation used in the study is part of the \emph{Sacado} package from the \emph{Trilinos} agile-component libraries developed at Sandia National Laboratories \cite{ref-Sacado}.

An AD system evaluates composition functions in the expected manner: it works by first evaluating the innermost function of the composition, then presenting that function's output as input to the next level function until all levels are complete, observing commonly expected order of operations. However, an AD system does additional work during a function evaluation in that as each nested function is evaluated, the function's partial derivative with respect to the designated variables of the given input is also calculated.  This is possible because the elementary math functions are hard-coded into the AD source-code along with their analytical derivatives, and linked by special instructions, so that when the elementary math functions are called upon for computation, their partial derivatives may be computed and stored in a sequence. The AD system then multiplies the final sequence of partial derivatives together to produce the exact equivalent to taking a partial derivative of the corresponding composition function with respect to a designated variable at a particular value. It is obvious, but bears mentioning, that the AD system could simply store one value for partial derivatives, modifying it as appropriate for every function evaluation rather than keeping a sequence. This is significant because modifying a single value rather than keeping a list of values corresponding to
every level of a composition function, which itself may not be the sole one needing to be evaluated, represents a savings in memory usage, which at large scales is an active concern. In the literature, the AD scheme described here is called \emph{forward automatic-differentiation}. Only forward AD will be covered here since it is the implementation used in \emph{Sacado} and therefore in this study; however, the reader is referred to the introduction section of \cite{ref-AD-methods} and its citations for further information on AD, and particularly \cite{ref-on-AD} which is foundational.  Appendix~\ref{sec:appendixB} includes an example that walks the reader through the process a computer uses to compute derivatives via AD.  Some things to note about AD are that no approximation of derivatives is being made because the analytical forms of the partials of the elementary math functions are defined alongside them. The accuracy of AD is then limited by the precision of the AD system's definition of the elementary math functions and their partial derivatives.

@@

In solid and structural computational mechanics the tangent-stiffness matrix is a linearization operator that describes the stiffness of a system in response to small displacements imposed upon the current configuration of the system.  Mathematically speaking, the tangent-stiffness represents the gradient of a high-dimensional energy surface that ``points'' in the minimum direction. While the terminology of solid mechanics is used here, it's important to note that these operators appear in other physical settings and are known by other names, e.g. a \emph{transmissibility matrix} in the context of a Poisson problem or, more generally, a \emph{Jacobian matrix} in mathematical optimization.

While, for simplicity, the examples shown in the following refer only to linear problems, tangent-stiffness matrices generally arise in the context of non-linear analysis where Newton's method or quasi-Newton's method's are used in the minimization of a given residual function.  In this context, the tangent-stiffness matrix can be thought of as the linearization of the system about about a particular configuration.  It has been shown \cite{hughes1978consistent,hughes1978unconditionally} that in order to preserve the quadratic convergence properties of the Newton's methods, it is necessary that this linearization is carried out in a manner that is \emph{consistent} with the algorithmic constraint equations used when computing the internal forces that arise due to deformations.  An example of these constraint equations would be the Kuhn-Tucker conditions \cite{simo1998} that are used in integration of a flow rule for plasticity modeling.  If the continuum tangent-moduli are used in place of the \emph{consistent} or \emph{algorithmic tangent moduli} in the solution of a non-linear solid mechanics problem, convergence may still be achieved, but not in the quadratic manner that makes Newton's method attractive for this class of problems.  In the setting of non-linear solid mechanics, there is a very small class of material model algorithms in which consistent tangent-stiffness operators can be derived analytically.  Perfect plasticity and plasticity with isotropic hardening used in combination with a general nearest-point projection or \emph{radial return algorithm} are a few examples.  Therefore, when general models (and certainly more complex ones) are implemented into a general purpose computational mechanics code, the tangent-stiffness operators are typically defined via a numerical approximation.

A general mathematical formula for a tangent-stiffness matrix, $K_{ij}$, can be expressed in indicial notation as 
%
\begin{equation} 
  K_{ij} = \left. \frac{\partial F_i}{\partial X_j}\right|_{\vec{X}_0},
\end{equation}
%
where $F_i$ is the $i^{\mbox{th}}$ component of the vector valued function, $X_j$ is the $j^{\mbox{th}}$ component of the vector argument to $F$, and a particular value of the vector, $\vec{X}_0$ is chosen as the linearization point. One then evaluates the expression for each combination $(i, j)$ corresponding to a (row, column) location in the tangent-stiffness matrix. The elements of a tangent-stiffness matrix can be estimated using any of the complex-step, AD, or common finite-difference techniques for functions $F:R^1 \rightarrow R^1$, since taking partial derivatives entails holding all but a single independent vector component of the function constant, and each element of the function can be evaluated independently of the others. \\

@@

The \emph{forward-difference} (FD) formula for calculating a tangent-stiffness matrix in the setting of a solid structural mechanics problem is 
%
\begin{equation} 
  K_{ij} = \frac{F_i^{int}(\vec{u} + h \hat{e}_j) - F_i^{int}(\vec{u})}{h},
\end{equation}
%
Where $K_{ij}$ is the indicial notation representation of an element of the tangent-stiffness matrix at row $i$, column $j$. The first of the two terms in the numerator is the internal force, $F_i^{int}$ evaluated in the current deformed configuration, $\vec{u}$,  plus a small perturbation $h$ in the direction $\hat{e}_j$ where $\hat{e}_j$ represents a unit-vector corresponding to the $j^{\mbox{th}}$ degree-of-freedom. The second of the two terms is $F_i^{int}$ evaluated in the current configuration. The denominator is the magnitude of perturbation called the \emph{probe-distance}.

The \emph{central-difference} (CD) formula for calculating a tangent-stiffness matrix is
%
\begin{equation} 
  K_{ij} = \frac{F_i^{int}(\vec{u} + h \hat{e}_j) - F_i^{int}(\vec{u} - h \hat{e}_j)}{2 h}.
\end{equation}
%
Together, the FD and CD methods, are sometimes termed \emph{finite-difference probing} techniques for tangent-stiffness calculation in literature \cite{ref-Adaggio}. From the well-known 1-dimensional formulas for FD and CD, it is understood that the accuracy of these expressions is dependent theoretically upon selecting a small step-size $h$, however experiments shown in \cite{squire1998using} demonstrate that too small an $h$ can lead to inaccuracy from ``subtractive-cancellation''. Therefore it is fair to say that a general method for selecting $h$ in practice is unobtainable.

The complex-step (CS) formula for calculating a tangent-stiffness matrix is
%
\begin{equation} 
  K_{ij} = \frac{\mbox{Im}(F_i^{int}(\vec{u} + i h \hat{e}_j))}{h},
\end{equation}
%
where the internal force, $F^{int}$ is now treated as function of complex vectors,  $\vec{u}$. The small perturbation, $h$, is now carried out on the imaginary axis and only the real coefficient to the imaginary part of the vector returned by $F^{int}$ is kept, hence the $\mbox{Im}(\cdot)$ operation. As highlighted in Section~\ref{sec:CSmethod}, notice that there is no subtraction operation taking place; therefore $h$ can be made very small to yield highly accurate derivatives. It should be mentioned that using complex-step to compute tangent-stiffness matrices was done in \cite{perez2000numerical,perez2012numerical}, but those references did not investigate the performance of complex-step in a massively parallel setting nor make comparisons to AD.

@@

\section{Description of analysis tools and approach.} 
%
\subsection{\emph{Peridigm} and  peridynamics.}
%
Comparisons of the different tangent-stiffness calculation techniques were carried out through code-development and \emph{in situ} instrumentation of the computational peridynamics code \emph{Peridigm} \cite{peridigm}. \emph{Peridigm} is distributed by Sandia National Laboratories as its primary open-source computational peridynamics code. It is a massively parallel simulation code for implicit and explicit multi-physics simulations primarily used for solid mechanics and material failure. \emph{Peridigm} is a C++ code utilizing agile software components from Sandia's  \emph{Trilinos} project \cite{trilinos}. It's important to note that there is no specialization of the tangent-stiffness calculation methods described previously to this particular code or more generally to a computational peridynamics approach and the analysis carried out in this study should provide insight into the speed and accuracy of these methods when implemented into other computational mechanics analysis tools as well.  \emph{Peridigm} was chosen primarily because the FD, CD, and AD methods for tangent-stiffness calculation where already implemented in the code, leaving only the CS method for development. Also, because peridynamics is a nonlocal theory, the tangent-stiffness matrices have a much higher bandwidth (i.e. less sparsity) than traditional finite-element tangent-stiffness computations and the time required to construct the tangent-stiffness is a majority of the total computation time in any quasi-static or implicit dynamics analysis; therefore, the \emph{Peridigm} development team has an interest in profiling the performance of the tangent-stiffness computation methods.

Briefly, peridynamics \cite{silling2000ret,silling:psa,silling2010peridynamic} is a nonlocal reformulation of the partial-differential equations that provide the statement of momentum balance in classical continuum mechanics. Its primary goal is to avoid the use of spatial derivatives in the balance equations or constitutive laws such that discontinuous displacements, i.e. cracks, are mathematically consistent with the governing equations. It has demonstrated great promise in modeling problems with pervasive failure \cite{littlewood2010}, interesting phenomenon such as dynamic crack branching \cite{ha2010sod}, and is being extended to multiphyics phenomena \cite{bobaru2011peridynamic,katiyar2013} and the problem of multi-scale coupling to molecular simulations \cite{seleson2009peridynamics}.

\subsubsection{Implementing the complex-step method in Peridigm} 
%
The CS method was implemented in \emph{Perdigm} in a manner that follows closely the implementation of the FD technique.  However, because of \emph{Peridigm}'s reliance on \emph{Trilinos} and specifically the \emph{Epetra} package, special steps had to be followed to allow the use of complex number data types.  This is because \emph{Epetra} vectors, which can be thought of as distributed memory parallel cousins to standard C++ STL vectors, are hard coded to be of type {\tt double} and can not be simply be declared as having a complex data type. Therefore, program methods that were used in the course of applying the FD method were copied, renamed and re-written to follow the CS formula. This involved changing the data type of intermediate variables which were locally scoped, and dynamically casting persistent memory variables to complex data types where necessary. No attempt was made to ``tune'' the CS code in any way for performance. The algorithm  was implemented in such as way that it was in one-to-one correspondence with the FD and CD methods; however, it must be noted that the overhead associated with the dynamic casts in the CS method likely hurts its overall performance and this issue could be alleviated by using the next generation of templated \emph{TPetra} vectors available in \emph{Trilinos} that are capable of taking an explicit instantiation of any data type including {\tt complex}. The version of \emph{Peridigm} modified to implement CS can be found in the paper repository.

\emph{Peridigm} was primarily written by computational scientists; therefore, many advanced techniques in C++ such as multiple virtual inheritance, pointer arithmetic, and distributed data structures were used in the basic code and by necessity in the modified code.  To see CS and other methods in a more narrow context, it is advised that readers also take a look at the simple examples provided on the corresponding author's website.
